Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-01-31 overview.tex
Streamlined section 2.1  More work required.

The usual Eulerian equations of gas dynamics with gravity are given by

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \vecv \cdot \nabla \rho
    & = & - \rho \nabla \cdot \vecv,
        \label{eq:mass} \\
%
\frac{\partial \vecv }{ \partial t} 
          + ( \vecv \cdot \nabla ) \vecv 
    & = & - \frac{1}{\rho} \nabla p
          - \nabla \phi,
        \label{eq:momentum} \\
%
 \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
          - \vecv \cdot \nabla \phi
%\nonumber          \\  & &     % uncomment for 2-column
     + \Gamma - \Lambda
        \label{eq:total_energy}
\end{eqnarray}
%
Here $E$, $e$, $\rho$, $\vecv$ and $p$ are the baryonic total specific fluid energy, the baryonic specific thermal energy, the baryonic density, velocity and pressure, respectively.  
%Here we use primes to indicate variables defined in the absolute frame of reference (to distinguish them from a comoving frame).  
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation. 
Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy_prime}).

@@

This set of coupled hydrodynamic plus gravity equations is the standard one that \enzo solves.  However, in the cosmological case it is useful to transform these to a coordinate system that is comoving with the expanding universe \citep{Peebles93} (but note that \enzo\ can also operate in a fixed coordinate system). Specifically, we define:
\begin{eqnarray}
\vecx & = & \vecr/a, \\
\vecvp & = & a \hspace{0.5mm} d\vecx/dt = 
              \vecv - \dot{a}\vecx, \\
\rhop    & = & a^3 \rho,   \\
\pp       & = & a^3 p, \\
\Ep       & = & E - 
              \dot{a} \vecx \cdot \vecvp - 
              \frac{1}{2} \dot{a}^2 \vecx^2, \\
\phip    & = & \phi + \frac{1}{2} a \ddot{a} \vecx^2.
\end{eqnarray}

These definitions result in the following comoving equivalents to eqs. 
(\ref{eq:mass})-(\ref{eq:potential}).
%

\begin{eqnarray} 
\frac{\partial \rhop}{\partial t} 
          + \frac{1}{a} \vecvp \cdot \nabla \rhop 
    & = & - \frac{1}{a} \rhop \nabla \cdot \vecvp, 
        \label{eq:mass_prime} \\
%
\frac{\partial \vecvp}{\partial t}  
          + \frac{1}{a} ( \vecvp \cdot \nabla ) \vecvp 
    & = & - \frac{\dot{a}}{a} \vecvp 
          - \frac{1}{a \rhop} \nabla \pp 
          - \frac{1}{a} \nabla \phip, 
        \label{eq:momentum_prime} \\
%
\frac{\partial \Ep}{\partial t} 
         + \frac{1}{a} \vecvp \cdot \nabla \Ep 
   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
         - \frac{1}{a \rhop } \nabla \cdot (p \vecvp) 
%\nonumber   & &   %uncomment for 2-column
         - \frac{1}{a} \vecvp \cdot \nabla \phip 
                + \Gamma - \Lambda.
       \label{eq:total_energy_prime} \\
%
\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
        \label{eq:total_energy_def_prime} \\
%
\ep                  & = & 
        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
      \label{eq:state_prime} \\ 
{\nabla}^2 \phip    & = & 
        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
      \label{eq:poisson}
\end{eqnarray}
%
%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%The density, however, is comoving:
%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of a smooth, homogeneous background, where $z$, the redshift, is a function only of $t$.  All derivatives are determined with respect to the comoving position $\vecx$, which is defined simply to remove the universal expansion from the coordinate system.

@@

It is clear that these equation are equivalent to the absolute coordinate version in the limit in which $a = 1$ and $\dot{a} = 0$.  In the rest of this paper we will employ only the unprimed variables with the understanding that we are not restricting ourselves to cosmological applications.

>>>>

The usual Eulerian equations of ideal magnetohydrodynamics with gravity in a coordinate systems comoving the cosmological expansion are given by

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \frac{1}{a} \vecv \cdot \nabla \rho
    & = & - \frac{1}{a} \rho \nabla \cdot \vecv, 
        \label{eq:mass} \\
%
\frac{\partial \vecv}{\partial t}  
          + \frac{1}{a} ( \vecv \cdot \nabla ) \vecv
    & = & - \frac{\dot{a}}{a} \vecv 
          - \frac{1}{a \rho} \nabla p
          - \frac{1}{a} \nabla \phi, 
        \label{eq:momentum} \\
%
\frac{\partial \Ep}{\partial t} 
         + \frac{1}{a} \vecv \cdot \nabla E
   & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
         - \frac{1}{a \rho } \nabla \cdot (p \vecv) 
%\nonumber   & &   %uncomment for 2-column
         - \frac{1}{a} \vecv \cdot \nabla \phi 
                + \Gamma - \Lambda.
       \label{eq:total_energy} 
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $e$, $\rho$, $\vecv$ and $p$ are the comoving total specific fluid energy, specific thermal energy, baryonic density, velocity and pressure, respectively.  
%Here we use primes to indicate variables defined in the absolute frame of reference (to distinguish them from a comoving frame).  
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation. 
Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy_prime}).

@@

For completeness, we note here that as part of the usual transformation to comoving coordinates, we have defined the following comoving quantities (primes indicate variables defined in the usual, proper, coordinate system).

This set of coupled hydrodynamic plus gravity equations is the standard one that \enzo solves.  However, in the cosmological case it is useful to transform these to a coordinate system that is comoving with the expanding universe \citep{Peebles93} (but note that \enzo\ can also operate in a fixed coordinate system). Specifically, we define:
\begin{eqnarray}
\vecx & = & \vecr/a, \\
\vecv & = & a \hspace{0.5mm} d\vecx/dt = 
              \vecvp - \dot{a}\vecx, \\
\rho    & = & a^3 \rhop,   \\
p       & = & a^3 \pp, \\
E       & = & \Ep - 
              \dot{a} \vecx \cdot \vecvp - 
              \frac{1}{2} \dot{a}^2 \vecx^2, \\
\phi    & = & \phip + \frac{1}{2} a \ddot{a} \vecx^2.
\end{eqnarray}

%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \nabla \rhop 
%    & = & - \frac{1}{a} \rhop \nabla \cdot \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \nabla ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \nabla \pp 
%          - \frac{1}{a} \nabla \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \nabla \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \nabla \cdot (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \nabla \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of a smooth, homogeneous background, where $z$, the redshift, is a function only of $t$.  All derivatives are determined with respect to the comoving position $\vecx$, which is defined simply to remove the universal expansion from the coordinate system.

@@

It is clear that the evolution equation are equivalent to the fixed coordinate version in the limit in which $a = 1$ and $\dot{a} = 0$.  We include the cosmological terms with the understanding that we are not restricting ourselves to cosmological applications.
